---
title: "Capstone: Shifting Tornado Hazard in the United States and Emerging Insurance Exposure"
author: "Jill_Baker"
date: "2026-03-21"
output: html_document
---

Capstone Project: Shifting Tornado Hazard in the United States and Emerging Insurance Exposure
Utilizing RStudio - Required Packages and Libraries below for install

#####################################
######## Installing packages ########
#####################################

```{r}
install.packages("ggplot2", repos = "https://cloud.r-project.org/")
options(repos = c(CRAN = "https://cloud.r-project.org/"))
install.packages("dplyr")
install.packages("tidycensus")
install.packages("PRROC")   
install.packages("randomForest")  
```
######################################
######## Installing libraries ########
######################################

```{r}
library(ggplot2)
library(dplyr)
library(tidymodels)
library(tidyr)
library(stringr)
library(tidycensus)
library(e1071)
library(reshape2)
library(caret)
library(PRROC)
library(randomForest)
library(glmnet)
library(xgboost)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

################
## Raw Data ####
################

## Read in the CSV for National Weather Service Tornado Data from 1950 - 2024 & view

```{r}
nsw_tornado_data <- read.csv("~/Documents/Merrimack_Data_Science/Capstone/Capstone_Datasets/Potential_Datasets/Dataset_1/1950-2024_all_tornadoes.csv", na.strings = "?", stringsAsFactors = T)
View(nsw_tornado_data)
```

# Quick view of the column names of the tornado dataset

```{r}
colnames(nsw_tornado_data)
```

## First clean to allow merging with ACS Data: 

## Merge column str with columns f1, f2, f3 and f4 to provide the full county FIPS number. Leading zeroes were dropped and need to add back in. FIPS full number will associate with a specific country where the tornado started, passed through and ended. Then merge stf plus f1, stf plus f2, etc. to form each FIPS Code. 

```{r}
nsw_tornado_data_clean <- nsw_tornado_data %>%
  mutate(
    stf = if_else(!is.na(stf), str_pad(as.character(stf), 2, pad = "0"), NA_character_),

    f1 = if_else(!is.na(f1) & f1 != 0, str_pad(as.character(f1), 3, pad = "0"), NA_character_),
    f2 = if_else(!is.na(f2) & f2 != 0, str_pad(as.character(f2), 3, pad = "0"), NA_character_),
    f3 = if_else(!is.na(f3) & f3 != 0, str_pad(as.character(f3), 3, pad = "0"), NA_character_),
    f4 = if_else(!is.na(f4) & f4 != 0, str_pad(as.character(f4), 3, pad = "0"), NA_character_),

    county_fips_1 = if_else(!is.na(f1), paste0(stf, f1), NA_character_),
    county_fips_2 = if_else(!is.na(f2), paste0(stf, f2), NA_character_),
    county_fips_3 = if_else(!is.na(f3), paste0(stf, f3), NA_character_),
    county_fips_4 = if_else(!is.na(f4), paste0(stf, f4), NA_character_)
  )

nsw_tornado_data_clean
```

## Loading Census data from American Community Survey (ACS) within the tidycensus package/library for population. 
## Will be able to merge datasets using common column for GEOID with FIPS in the tornado dataset. 

#Viewing which variables to pull for population 

```{r}
acs_vars <- load_variables(2024, "acs5", cache = TRUE)
acs_vars
```

```{r}
population_variables <- acs_vars %>%
  filter(str_detect(concept, "Population"))
population_variables 
```

## Selecting B01003_001 for Total Population Estimate

## Reviewing variables for housing data in the census 

```{r}
housing_variables <- acs_vars %>%
  filter(str_detect(concept, "Housing"))
housing_variables 
```

## Selecting B25001_001 for Total Housing Units 

## Reviewing variables for home value in the census next:

```{r}
housing_variables <- acs_vars %>%
  filter(str_detect(concept, "Value"))
housing_variables 
```

## Selecting B25077_001 for Median Home Value & writing CSV for population for GitHub

```{r}
population <- get_acs(
  geography = "county",
  variables = "B01003_001",  
  year = 2024
)
population 

write.csv(population, "acs_county_population_2024.csv", row.names = FALSE)
getwd()
```

## Merge data on county_fips_f1 only. This country id is where the tornado originated and will provide sufficient data on risk. I do not want to merge data on all 4 FIPS (which just shows the counties the tornado passed through) to avoid duplicate data. This will still produce the end result of showing high-risk counties.

###############################
##### MERGE AND CLEAN DATA ####
###############################

## Cleaning Population dataset to rename GEOID to county_fips and estimate to unique population_estimate

```{r}
population_clean <- population %>%
  select(GEOID, NAME, estimate) %>%
  rename(
    county_fips = GEOID,
    population_estimate = estimate
  )

population_clean
```

## Cleaned tornado data with county_fips

```{r}
tornado_f1_fips <- nsw_tornado_data_clean %>%
  select(om, yr, mo, dy, date, st, stf, mag, inj, loss, len, wid, county_fips_1) %>%
  filter(!is.na(county_fips_1)) %>%
  rename(county_fips = county_fips_1)

tornado_f1_fips
```

## Merging tornado data with census population data on column county_fips

```{r}
tornado_merged_census <- tornado_f1_fips %>%
  left_join(population_clean, by = "county_fips")

tornado_merged_census
```

## To merge with Estimate Total Housing Units and Median Value & writing CSV files for Housing Units and Median Value for GitHub

```{r}
housing_units <- get_acs(
  geography = "county",
  variables = "B25001_001",  
  year = 2024
)
housing_units 

write.csv(housing_units, "acs_county_housing_units_2024.csv", row.names = FALSE)
```
```{r}
median_value <- get_acs(
  geography = "county",
  variables = "B25077_001",  
  year = 2024
)
median_value
write.csv(median_value, "acs_county_median_value_2024.csv", row.names = FALSE)
```

## Cleaning to rename column GEOID to county_fips for merging and specify the estimate column for unique values in Housing Units an Median Value: 

```{r}
housing_clean <- housing_units %>%
  select(GEOID, estimate, moe) %>%
  rename(
    county_fips = GEOID, 
    housing_unit_estimate = estimate,
    housing_moe = moe
  )

housing_clean
```

```{r}
medvalue_clean <- median_value %>%
  select(GEOID, estimate, moe) %>%
  rename(
    county_fips = GEOID,
    medvalue_estimate = estimate,
    medvalue_moe = moe
  )

medvalue_clean
```
################################
###### FINAL RAW DATA SET ######
################################

## Final merge -- all datasets merged: tornado, population, housing and median value. 
## This is the FINAL RAW DATASET BEFORE FEATURE ENGINEERING - write.csv for GitHub

```{r}
final_raw_dataset <- tornado_merged_census %>%
  left_join(housing_clean, by = "county_fips") %>%
  left_join(medvalue_clean, by = "county_fips")

final_raw_dataset
write.csv(final_raw_dataset, "final_raw_dataset.csv", row.names = FALSE)
```




#######################################################################################################################


#########################################
###### Feature Engineering Begins #######
#########################################

#Grouping Tornado Alley and Dixie Alley states

```{r}
tornado_alley_states = c("TX", "OK", "KS", "NE")
dixie_alley_states = c("LA", "AR", "TN", "GA", "AL", "MS")
```

## Cleaning dataset to remove columns not needed: stf, inj, housing_moe, medvalue_moe
## Renaming column NAME to county_name_state

```{r}
cleaned_tornado_dataset <- final_raw_dataset %>%
  select(-stf, -inj, -housing_moe, -medvalue_moe) %>%
  rename(county_name_state = NAME)
```

## Feature Creation/Transformation for: region_category, tornado_count_county_year, average_intensity_county, max_intensity_county, total_property_loss_year, average_property_loss, strong_tornado_count, property_loss_per_tornado

#region_category: T for Tornado Alley State, D for Dixie Alley State, O for Other

```{r}
names(cleaned_tornado_dataset)
```

# Region Category 

```{r}
final_dataset_regions <- cleaned_tornado_dataset %>%
  mutate(
    region_category = case_when(
      st %in% tornado_alley_states ~ "T",
      st %in% dixie_alley_states ~ "D",
      TRUE ~ "O"
    ),
    region_category = factor(region_category, levels = c("O", "T", "D")))

final_dataset_regions
```

## tornado_count_county_year feature creation. Total number of tornadoes per county for a given year

```{r}
tornado_hazard_dataset <- final_dataset_regions %>%
  group_by(county_fips, yr) %>%
  mutate(
    tornado_count_county_year = n()
  ) %>%
  ungroup()
  
tornado_hazard_dataset
```

## Filtered data for years 1979-2024 for project 

```{r}
tornado_hazard_dataset_filtered <- tornado_hazard_dataset %>%
  filter(yr >=1979)

tornado_hazard_dataset_filtered
```

#Quick Analysis: Counties with most tornadoes

```{r}
tornado_hazard_dataset_filtered %>%
  filter(tornado_count_county_year == max(tornado_count_county_year, na.rm = TRUE))
```

#Quick Analysis: Tornado count per year

```{r}
tornadoes_by_year <- tornado_hazard_dataset_filtered %>%
  count(yr, name = "tornado_count")

tornadoes_by_year

ggplot(tornadoes_by_year, aes(x = yr, y = tornado_count)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  labs(
    title = "Tornado Count Per Year (1979–2024)",
    x = "Year",
    y = "Number of Tornadoes"
  ) +
  theme_minimal()
```

#Quick Analysis: Dixie Alley vs Tornado Alley frequency per year 

```{r}
tornadoes_DT <- tornado_hazard_dataset_filtered %>%
  filter(region_category %in% c("D", "T")) %>%
  count(yr, region_category, name = "tornado_count") %>%
  mutate(region_name = case_when(
    region_category == "D" ~ "Dixie Alley",
    region_category == "T" ~ "Tornado Alley"
   
    ))

ggplot(tornadoes_DT , aes(x = yr, y = tornado_count, color = region_name)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  labs(
    title = "Tornado Count Dixie Alley vs. Tornado Alley",
    x = "Year",
    y = "Number of Tornadoes",
    color = "Region"
  ) +
  theme_minimal()
```

```{r}
tornado_hazard_dataset_filtered
colnames(tornado_hazard_dataset_filtered)
```

### ENGINEERED COUNTY-YEAR MODELING DATASET 

## average_intensity_county, max_intenstiy, strong_tornado_count feature creation. Standarize mag - new column with NA

```{r}
unique(tornado_hazard_dataset_filtered$loss)
```

## Cleaning: Adjusting property loss for consistency - prior to 1996 and after. Prior to 1996 was categorical - finding midpoint for millions of dollars amount

```{r}
tornado_hazard_dataset_filtered <- tornado_hazard_dataset_filtered %>%
  mutate(
    loss = as.numeric(loss),
    loss_clean = case_when(
      is.na(loss) | loss == 0 ~ NA_real_,

      yr < 1996 & loss == 1 ~ 0.000025,
      yr < 1996 & loss == 2 ~ 0.000275,
      yr < 1996 & loss == 3 ~ 0.00275,
      yr < 1996 & loss == 4 ~ 0.0275,
      yr < 1996 & loss == 5 ~ 0.275,
      yr < 1996 & loss == 6 ~ 2.75,
      yr < 1996 & loss == 7 ~ 27.5,
      yr < 1996 & loss == 8 ~ 275,
      yr < 1996 & loss == 9 ~ 500,  

      yr >= 1996 ~ as.numeric(loss)
    )
  )
```


```{r}
summary(tornado_hazard_dataset_filtered$loss_clean)
sum(tornado_hazard_dataset_filtered$loss_clean == 0, na.rm = TRUE)
```

```{r}
tornado_hazard_dataset_filtered
```

## Cleaning tornado intensity (mag)
## mag_standardized to not allow for skewing of averages/max with -9 (unknown). Updating to NA value.

```{r}
tornado_hazard_dataset_filtered <- tornado_hazard_dataset_filtered %>%
  mutate(mag_standardized = ifelse(mag == -9, NA, mag))
tornado_hazard_dataset_filtered
```

## Tornado-level to county-year level data
## Feature Creation: Grouping by county_fips and year for county-year level data: 
## tornado_count_county_year - number of tornadoes that occurred
## average_intensity_county - mean/mag_standardized for average tornado intensity per county_year
## max_intensity_county - max/mag_standardized for max tornado intensity per county_year
## strong_tornado_count - sum of the number of mag_standardized tornadoes greater than an F/EF3
## total_property_loss_year - sum of the cleaned loss data per county_year
## average_property_loss - mean of the cleaned loss data per county_year
## avg_length - mean of the length of tornadoes per county_year
## avg_width - mean of the width of tornadoes per county_year

```{r}
county_year_hazard <- tornado_hazard_dataset_filtered %>%
  group_by(county_fips, yr) %>%
  summarise(
    tornado_count_county_year = n(),
    average_intensity_county = ifelse(all(is.na(mag_standardized)), NA_real_,
      mean(mag_standardized, na.rm = TRUE)),
    max_intensity_county = ifelse(all(is.na(mag_standardized)), NA_real_,
      max(mag_standardized, na.rm = TRUE)),
    strong_tornado_count = sum(mag_standardized >= 3, na.rm = TRUE),
    total_property_loss_year = ifelse(all(is.na(loss_clean)), NA_real_,
      sum(loss_clean, na.rm = TRUE)),
    average_property_loss = ifelse(all(is.na(loss_clean)), NA_real_,
      mean(loss_clean, na.rm = TRUE)),
    avg_length = ifelse(all(is.na(len)), NA_real_, mean(len, na.rm = TRUE)),
    avg_width = ifelse(all(is.na(wid)), NA_real_, mean(wid, na.rm = TRUE)),
    .groups = "drop"
  )
  
county_year_hazard 
```

## Accidentally ommited the county_names column in the dataset - readding it here

```{r}
county_names <- tornado_hazard_dataset_filtered %>%
  select(county_fips, county_name_state) %>%
  distinct()

county_year_hazard <- county_year_hazard %>%
  left_join(county_names, by = "county_fips")
county_year_hazard
```

## Create property loss per tornado feature by dividing total property loss per year by the total tornado count per county year:

```{r}
county_year_hazard <- county_year_hazard %>%
  mutate(
    property_loss_per_tornado = total_property_loss_year / tornado_count_county_year
  )

county_year_hazard
```


## Merge in the ACS county-level variables - ensuring tornado_alley_fips and dixie_alley_fips coordinate with the designated state fips codes: 

```{r}
tornado_alley_fips <- c("48", "40", "20", "31")  # TX, OK, KS, NE
dixie_alley_fips  <- c("22", "05", "47", "13", "01", "28")  # LA, AR, TN, GA, AL, MS
```

## Adding in population estimate, housing unit estimate and medvalue estimate values 

```{r}
county_lookup <- tornado_hazard_dataset_filtered %>%
  group_by(county_fips) %>%
  summarise(
    population_estimate = first(na.omit(population_estimate)),
    housing_unit_estimate = first(na.omit(housing_unit_estimate)),
    medvalue_estimate = first(na.omit(medvalue_estimate)),
    .groups = "drop"
  ) %>%
  mutate(
    state_fips = substr(county_fips, 1, 2),
    region_category = case_when(
      state_fips %in% tornado_alley_fips ~ "T",
      state_fips %in% dixie_alley_fips ~ "D",
      TRUE ~ "O"
    ),
    region_category = factor(region_category, levels = c("O", "T", "D"))
  )
```


```{r}
county_lookup %>%
  count(county_fips) %>%
  filter(n > 1)
```

## Census data now included in the tornado dataset with engineered features including region grouping 

```{r}
county_year_hazard <- county_year_hazard %>%
  left_join(
    county_lookup %>%
      select(county_fips, population_estimate, housing_unit_estimate, medvalue_estimate, region_category),
    by = "county_fips"
  )

colnames(county_year_hazard)
```

## Creating normalized risk 
## property_loss_per_tornado if any is 0 then value is NA ensuring no division by 0
## tornadoes_per_100k_population as population is vastly different across these regions this normalizes it to adjust tornado frequency by population size 
## tornadoes_per_10k_housing_units also adjusts tornado frequency by housing concentration 

```{r}
county_year_hazard <- county_year_hazard %>%
  mutate(
    property_loss_per_tornado = ifelse(
      is.na(total_property_loss_year) | tornado_count_county_year == 0,
      NA,
      total_property_loss_year / tornado_count_county_year
    ),
    tornadoes_per_100k_population = ifelse(
      is.na(population_estimate) | population_estimate == 0,
      NA,
      (tornado_count_county_year / population_estimate) * 100000
    ),
    tornadoes_per_10k_housing_units = ifelse(
      is.na(housing_unit_estimate) | housing_unit_estimate == 0,
      NA,
      (tornado_count_county_year / housing_unit_estimate) * 10000
    )
    
  )

colnames(county_year_hazard)
```
#################################
#### TARGET VARIABLE CREATION ###
#################################


##### Target variable creation: high_tornado_hazard_next_year ####

## Log transformation due to highly skewed property loss 

```{r}
county_year_hazard <- county_year_hazard %>%
  mutate(
    log_property_loss = log1p(total_property_loss_year)
  )
```

## Z-scores for target variable's hazard variables to allow standardization then create a composite hazard score with an average of all standardized variables: tornado_count_county_year, average_intensity_county, max_intensity_county, strong_tornado_count, log_property_loss 

```{r}
county_year_hazard <- county_year_hazard %>%
  mutate(
    z_tornado_count = as.numeric(scale(tornado_count_county_year)),
    z_avg_intensity = as.numeric(scale(average_intensity_county)),
    z_max_intensity = as.numeric(scale(max_intensity_county)),
    z_strong_count = as.numeric(scale(strong_tornado_count)),
    z_property_loss = as.numeric(scale(log_property_loss))
  ) %>%
  mutate(
    composite_hazard_score = rowMeans(
      cbind(
        z_tornado_count,
        z_avg_intensity,
        z_max_intensity,
        z_strong_count,
        z_property_loss
      ),
      na.rm = TRUE
    )
  )

county_year_hazard
```

## Defining the hazard cutoff for binary classification. 
## This was originally set at 75%, however, it was adjusted to 65% to allow for better performance (aka the top 35% of county-years will be classified as high-risk)
## Binary classification of 1 = high-risk and 0 = not high-risk 

```{r}
hazard_cutoff <- quantile(county_year_hazard$composite_hazard_score, probs = 0.65, na.rm = TRUE)

county_year_hazard <- county_year_hazard %>%
  mutate(
    high_tornado_hazard = ifelse(composite_hazard_score >= hazard_cutoff, 1, 0)
  )
```

## For the predicitve modeling aspect - there must be a shift forward one year 

```{r}
county_year_hazard <- county_year_hazard %>%
  arrange(county_fips, yr) %>%
  group_by(county_fips) %>%
  mutate(
    high_tornado_hazard_next_year = lead(high_tornado_hazard)
  ) %>%
  ungroup()
```

```{r}
county_year_hazard <- county_year_hazard %>%
  mutate(
    high_tornado_hazard = as.integer(high_tornado_hazard),
    high_tornado_hazard_next_year = as.integer(high_tornado_hazard_next_year)
  )
```

```{r}
table(county_year_hazard$high_tornado_hazard)
table(county_year_hazard$high_tornado_hazard_next_year)
summary(county_year_hazard$composite_hazard_score)
```

## Cleaned dataset for GitHub

```{r}
county_year_hazard
write.csv(county_year_hazard, "adjusted_county_year_hazard_model_dataset3.csv", row.names = FALSE)
```





#########################################
##### Exploratory Data Analysis/EDA #####
#########################################

## Viewing the min, max, mean and median of each continuous variable 

```{r}
summary(county_year_hazard)
```

## Log transformation highly right-skewed variables: population and housing unit 

```{r}
county_year_hazard <- county_year_hazard %>%
  mutate(
    log_population = log1p(population_estimate),
    log_housing = log1p(housing_unit_estimate)
  )
county_year_hazard
```

## Continuous Variable EDA - Viewing for mean, median, standard deviation, min, max and skewness for each variable: 

```{r}
continuous_vars <- county_year_hazard %>%
  select(
    population_estimate,
    housing_unit_estimate,
    medvalue_estimate,
    tornado_count_county_year,
    average_intensity_county,
    max_intensity_county,
    total_property_loss_year,
    average_property_loss,
    strong_tornado_count,
    property_loss_per_tornado,
    tornadoes_per_100k_population,
    tornadoes_per_10k_housing_units,
    avg_length,
    avg_width,
    composite_hazard_score,
    log_property_loss,
    log_population,
    log_housing
  )

continuous_summary_table <- data.frame(
  Variable = names(continuous_vars),
  Mean = sapply(continuous_vars, mean, na.rm = TRUE),
  Median = sapply(continuous_vars, median, na.rm = TRUE),
  Standard_Deviation = sapply(continuous_vars, sd, na.rm = TRUE),
  Min = sapply(continuous_vars, min, na.rm = TRUE),
  Max = sapply(continuous_vars, max, na.rm = TRUE),
  Skewness = sapply(continuous_vars, skewness, na.rm = TRUE)
)

continuous_summary_table
```

## EDA - Viewing count and percent for categorical variables 

```{r}
categorical_summary <- county_year_hazard %>%
  select(region_category, high_tornado_hazard, high_tornado_hazard_next_year) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Category") %>%
  group_by(Variable, Category) %>%
  summarise(
    Count = n(),
    .groups = "drop"
  ) %>%
  group_by(Variable) %>%
  mutate(
    Percent = round(Count / sum(Count) * 100, 2)
  ) %>%
  arrange(Variable, Category)

categorical_summary
```

#New Feature: Calculating the fraction of strong tornadoes out of the total tornado count
```{r}
county_year_hazard <- county_year_hazard %>%
  mutate(
    strong_tornado_fraction = ifelse(
      tornado_count_county_year == 0,
      NA,
      strong_tornado_count / tornado_count_county_year
    )
  )
```
#########################
####### EDA Plots #######
#########################

## Bivariate plot for Total Tornado Count by Year (1979-2024)
## Indicates an increase in total tornadoes over the years 

```{r}
tornadoes_by_year <- county_year_hazard %>%
  group_by(yr) %>%
  summarise(
    tornado_count = sum(tornado_count_county_year, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(tornadoes_by_year, aes(x = yr, y = tornado_count)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_smooth(se = FALSE, color = "darkorange", linewidth = 1.2) +
  labs(
    title = "Total Tornado Count by Year (1979–2024)",
    x = "Year",
    y = "Total Tornado Count",
    caption = "Annual tornado counts in the United States aggregated by year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    legend.position = "top",
    plot.caption = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 13, face = "bold"),
    axis.title.y = element_text(size = 13, face = "bold")
  )
```

## Plot for Total Tornado Count by County Year by Region (Tornado Alley vs. Dixie Alley 1979-2024)
## Indicates overall tornadoes counts have been increasing the the Dixie Alley over the years 

```{r}
region_year_counts <- county_year_hazard %>%
  filter(region_category %in% c("T", "D")) %>%
  group_by(yr, region_category) %>%
  summarise(
    tornado_count = sum(tornado_count_county_year, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    region_category = recode(region_category,
      "T" = "Tornado Alley",
      "D" = "Dixie Alley"
    )
  )

ggplot(region_year_counts, aes(x = yr, y = tornado_count, color = region_category)) +
  geom_line(linewidth = 0.5) +
  geom_smooth(se = FALSE, linewidth = 0.5, linetype = "dashed") +
  scale_color_manual(values = c("Tornado Alley" = "steelblue", "Dixie Alley" = "darkorange")) +
  labs(
    title = "Total Tornado Count by Year in Tornado Alley vs. Dixie Alley (1979–2024)",
    x = "Year",
    y = "Total Tornado Count",
    color = "Region",
    caption = "Annual tornado counts by year in Tornado Alley & Dixie Alley"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 13, face = "bold"),
    axis.title.y = element_text(size = 13, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "top",
    legend.justification = "center",
    legend.box.just = "center",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 11)
  )
```

## Box Plot - Tornadoes per 100,000 Population by Region (Tornado Alley & Dixie Alley)
## Indicates higher number of outliers within Tornado Alley 

```{r}
ggplot(
  county_year_hazard %>% 
    filter(region_category %in% c("T", "D")),
  aes(
    x = tornadoes_per_100k_population,
    fill = region_category,
    color = region_category
  )
) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("T" = "blue", "D" = "orange")) +
  scale_color_manual(values = c("T" = "blue", "D" = "orange")) +
  labs(
    title = "Distribution of Tornadoes per 100,000 Population by Region",
    x = "Tornadoes per 100,000 Population",
    y = "Density",
    caption = "Distribution comparison between Tornado Alley and Dixie Alley"
  ) +
   scale_x_log10()+
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 11, face = "bold"),
    axis.title.y = element_text(size = 11, face = "bold"),
    legend.title = element_blank()
  )
```

## Box Plot - Tornadoes per 10,000 Housing Units by Region
## Also indicates a higher number of outliers within Tornado Alley 

```{r}
ggplot(
  county_year_hazard %>% 
    filter(region_category %in% c("T", "D")),
  aes(
    x = tornadoes_per_10k_housing_units,
    fill = region_category,
    color = region_category
  )
) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(
    values = c("T" = "blue", "D" = "orange"),
    labels = c("T" = "Tornado Alley", "D" = "Dixie Alley")
  ) +
  scale_color_manual(
    values = c("T" = "blue", "D" = "orange"),
    labels = c("T" = "Tornado Alley", "D" = "Dixie Alley")
  ) +
  scale_x_log10() +
  labs(
    title = "Distribution of Tornadoes per 10,000 Housing Units by Region",
    x = "Tornadoes per 10,000 Housing Units",
    y = "Density",
    caption = "Distribution comparison between Tornado Alley and Dixie Alley"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 11, face = "bold"),
    axis.title.y = element_text(size = 11, face = "bold"),
    legend.position = "right"
  )
```

## Plot of Strong Tornado Count by Year in Tornado Alley vs. Dixie Alley (1979–2024)

```{r}
strong_by_year <- county_year_hazard %>%
  filter(region_category %in% c("T", "D")) %>%
  group_by(yr, region_category) %>%
  summarise(
    total_strong_tornadoes = sum(strong_tornado_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    region_category = recode(region_category,
      "T" = "Tornado Alley",
      "D" = "Dixie Alley"
    )
  )

ggplot(strong_by_year, aes(x = yr, y = total_strong_tornadoes, color = region_category)) +
  geom_line(linewidth = 0.5) +
  geom_smooth(se = FALSE, linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Tornado Alley" = "steelblue", "Dixie Alley" = "darkorange")) +
  labs(
    title = "Strong Tornado Count by Year in Tornado Alley vs. Dixie Alley (1979–2024)",
    x = "Year",
    y = "Total Strong Tornado Count (F/EF Scale 3+)",
    color = "Region",
    caption = "Annual strong tornado counts by region, Tornado Alley and Dixie Alley, (1979–2024)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 13, face = "bold"),
    axis.title.y = element_text(size = 13, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 11)
  )
```

#New Graph: Fraction of strong tornadoes dividied by total tornadoes per county year 

```{r}
ggplot(
  county_year_hazard %>%
    filter(region_category %in% c("T", "D")),
  aes(x = yr, y = strong_tornado_fraction, color = region_category)
) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  scale_color_manual(
    values = c("T" = "steelblue", "D" = "darkorange"),
    labels = c("T" = "Tornado Alley", "D" = "Dixie Alley")
  ) +
  labs(
    title = "Average Proportion of Strong Tornadoes Over Time",
    x = "Year",
    y = "Proportion of Strong Tornadoes",
    color = "Region",
    caption = "Average proprotion of strong tornadoes (F/EF3 or greater) over time"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 13, face = "bold"),
    axis.title.y = element_text(size = 13, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 11)
  )
```

## Correlation matrix of continuous numeric variables included in the analysis
## Indicating some multicollinearity with population and housing units 

```{r}

correlation_data <- county_year_hazard %>%
  select(
    population_estimate,
    housing_unit_estimate,
    medvalue_estimate,
    tornado_count_county_year,
    average_intensity_county,
    max_intensity_county,
    total_property_loss_year,
    average_property_loss,
    strong_tornado_count,
    property_loss_per_tornado,
    tornadoes_per_100k_population,
    tornadoes_per_10k_housing_units,
    avg_width,
    avg_length
  )

correlation_matrix <- cor(correlation_data, use = "complete.obs")

correlation_melt <- melt(correlation_matrix)

ggplot(correlation_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "darkorange",
    mid = "white",
    high = "steelblue",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation Heatmap of Continuous Numeric Variables",
    x = NULL,
    y = NULL,
    fill = "Correlation",
    caption = "A correlation matrix of continuous numeric variables included in the analysis"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.3, face = "bold"),
    plot.caption = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9, face = "bold"),
    axis.text.y = element_text(size = 9, face = "bold")
  )
```

## Plots for visualization of improvements of log transformations for population, housing units and property loss (before and after log transformations)

```{r}
## Population

population_long <- county_year_hazard %>%
  select(population_estimate, log_population) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(population_long, aes(x = value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", na.rm = TRUE) +
  facet_wrap(~variable, scales = "free") +
  labs(
    title = "Population Distribution Before and After Log Transformation",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

## Housing Units 

housing_long <- county_year_hazard %>%
  select(housing_unit_estimate, log_housing) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(housing_long, aes(x = value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", na.rm = TRUE) +
  facet_wrap(~variable, scales = "free") +
  labs(
    title = "Housing Unit Distribution Before and After Log Transformation",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

## Property Loss 

property_long <- county_year_hazard %>%
  select(total_property_loss_year, log_property_loss) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(property_long, aes(x = value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", na.rm = TRUE) +
  facet_wrap(~variable, scales = "free") +
  labs(
    title = "Property Loss Distribution Before and After Log Transformation",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```
###############################################################################

##################
#### Modeling ####
##################

## Indicating labels for the target variables 0 = No Risk and 1 = High Risk

## Labeling the high tornado hazard next year target - 0 = no risk ,1 = high risk

```{r}
county_year_hazard <- county_year_hazard %>%
  mutate(
    high_tornado_hazard_next_year_factor = factor(
      high_tornado_hazard_next_year,
      levels = c(0, 1),
      labels = c("NoRisk", "HighRisk")
    )
  )
```


```{r}
table(county_year_hazard$high_tornado_hazard_next_year_factor, useNA = "always")
```

## Modeling Dataset and Split 
##Separate train/test split time-based. Train is anything from 1979-2014 and test is anything from 2014-2025

```{r}
predictive_model_data <- county_year_hazard %>%
  select(
    yr,
    county_fips,
    county_name_state,
    high_tornado_hazard_next_year_factor,
    population_estimate,
    housing_unit_estimate,
    medvalue_estimate,
    region_category,
    avg_length,
    avg_width
  ) %>%
  filter(!is.na(high_tornado_hazard_next_year_factor))
```

## Filtered Train and Test Splits 

```{r}
train_set <- predictive_model_data %>%
  filter(yr >= 1979 & yr <= 2014) %>%
  select(-yr)

test_set <- predictive_model_data %>%
  filter(yr >= 2015 & yr <= 2024) %>%
  select(-yr)

set.seed(123)
```

## Model Train and Test Sets with predictor variables selected (target variable plus population, housing, median value, region, average length and average width)

```{r}
model_train <- train_set %>%
  select(
    high_tornado_hazard_next_year_factor,
    population_estimate,
    housing_unit_estimate,
    medvalue_estimate,
    region_category,
    avg_length,
    avg_width
  )

model_test_full <- test_set

model_test <- test_set %>%
  select(
    high_tornado_hazard_next_year_factor,
    population_estimate,
    housing_unit_estimate,
    medvalue_estimate,
    region_category,
    avg_length,
    avg_width
  )
```


#########################################
#### PCA Dimensionality Reduction #######
#########################################

```{r}
pca_data <- county_year_hazard %>%
  select(
    population_estimate,
    housing_unit_estimate,
    medvalue_estimate,
    avg_length,
    avg_width,
    tornadoes_per_100k_population,
    tornadoes_per_10k_housing_units
  ) %>%
  na.omit()

pca_scaled <- scale(pca_data)

pca_model <- prcomp(pca_scaled, center = TRUE, scale. = TRUE)

summary(pca_model)

```

####################################
### Feature Importance Selection ###
####################################

## Logistic Regression: Coefficient Magnitude with L1 Regularization (Lasso) 
## random forest: gini
## XG Boost: gain

## Lasso 

```{r}
nrow(predictive_model_data)
```

## Remove NA values in predictors (received an error without doing so)

```{r}
lasso_data <- model_train %>%
  na.omit()

logistic_regression_x_train <- model.matrix(high_tornado_hazard_next_year_factor ~ ., lasso_data)[, -1]
logistic_regression_y_train <- as.numeric(lasso_data$high_tornado_hazard_next_year_factor) - 1

feature_importance_lasso_model <- cv.glmnet(logistic_regression_x_train, logistic_regression_y_train, family = "binomial", alpha = 1)

coef(feature_importance_lasso_model, s = "lambda.min")
```

## Quick interpretation of results from Lasso: 
## population estimate zeroed
## positive importance features: Region category (Dixie Alley), average legnth, average width, region category (Tornado Alley), housing unit and negative median value 

#Random Forest - gini

```{r}
rf_data <- model_train %>%
  na.omit()

set.seed(123)

feature_importance_random_forest_model <- randomForest(
  high_tornado_hazard_next_year_factor ~ .,
  data = rf_data, 
  importance = TRUE
)

importance(feature_importance_random_forest_model)
varImpPlot(feature_importance_random_forest_model)
```

## Quick Interpretations of random forest using mean decrease gini with highest to lowest importance: average length, median value, population, housing unit, average width, region category. Random forest indicates that region category is lowest importance. 

#XG Boost - gain 

```{r}

xgb_data <- model_train %>%
  na.omit()

xgboost_x_train <- model.matrix(high_tornado_hazard_next_year_factor ~ ., xgb_data)[, -1]
xgboost_y_train <- as.numeric(xgb_data$high_tornado_hazard_next_year_factor) - 1

set.seed(123)
feature_importance_xgboost_model <- xgboost(
  data = xgboost_x_train, 
  label = xgboost_y_train,
  nrounds = 100, 
  objective = "binary:logistic",
  verbose = 0
)

xgb_importance <- xgb.importance(model = feature_importance_xgboost_model)
xgb_importance

```

## Quick interpretation of XG Boost: highest gain is average length and median value estimate, followed by population, average width and housing unit. Least gain/importance is region: Dixie and Tornado Alley

## Comparison of the 3 models - creation of dataframes for each: 

## Lasso 

```{r}
lasso_df <- data.frame(
  Feature = rownames(coef(feature_importance_lasso_model, s = "lambda.min")),
  Importance = as.numeric(coef(feature_importance_lasso_model, s = "lambda.min"))
) %>%
  filter(Feature != "(Intercept)") %>%
  mutate(Model = "LASSO")
```

## random forest

```{r}
rf_df <- data.frame(
  Feature = rownames(importance(feature_importance_random_forest_model)),
  Importance = importance(feature_importance_random_forest_model)[, "MeanDecreaseGini"]
) %>%
  mutate(Model = "Random Forest")
```

## XG Boost 

```{r}
xgb_df <- xgb_importance %>%
  select(Feature, Gain) %>%
  rename(Importance = Gain) %>%
  mutate(Model = "XGBoost")
```

## Combining all 3 & normalizing (as they are all different scales)

```{r}
importance_all <- bind_rows(lasso_df, rf_df, xgb_df)

importance_all <- importance_all %>%
  group_by(Model) %>%
  mutate(Importance_scaled = abs(Importance) / max(abs(Importance))) %>%
  ungroup()

importance_all <- importance_all %>%
  group_by(Feature) %>%
  mutate(avg_importance = mean(Importance_scaled)) %>%
  ungroup()

importance_all <- importance_all %>%
  arrange(desc(avg_importance)) %>%
  mutate(Feature = factor(Feature, levels = unique(Feature)))
```

## Plot to compare feature importance across each model: logistic regression, random forest and XG Boost 

```{r}
ggplot(importance_all, aes(x = Feature, 
                           y = Importance_scaled, 
                           fill = Model)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  coord_flip() +
  scale_fill_manual(values = c(
    "LASSO" = "grey",
    "Random Forest" = "steelblue",
    "XGBoost" = "darkorange"
  )) +
  labs(
    title = "Feature Importance Comparison Across Models",
    x = "Feature",
    y = "Normalized Importance (0–1)"
  ) +
  theme_minimal(base_size = 12)
```

###################################
######### MODEL EVALUATION ########
###################################

## custom function for consistent classification thresholding across models
```{r}
get_predictions <- function(probs, threshold = 0.3) {
  preds <- ifelse(probs > threshold, "HighRisk", "NoRisk")
  return(factor(preds, levels = c("NoRisk", "HighRisk")))
}

```

## custom function for PR-AUC instead of ROC (with caret

```{r}
prSummary <- function(data, lev = NULL, model = NULL) {
  require(PRROC)
  
  if (length(unique(data$obs)) < 2) {
    return(c(PR_AUC = NA))
  }
  
  positive_class <- lev[2]
  
  scores <- data[[positive_class]]
  labels <- ifelse(data$obs == positive_class, 1, 0)
  
  pr <- pr.curve(
    scores.class0 = scores[labels == 1],
    scores.class1 = scores[labels == 0],
    curve = FALSE
  )
  
  c(PR_AUC = pr$auc.integral)
}
```


#Cross Validation & tuning on Training Set ONLY 
## Stratified k-fold

```{r}
model_train_clean <- model_train %>%
  na.omit()

model_train_clean$strata <- interaction(
  model_train_clean$high_tornado_hazard_next_year_factor,
  model_train_clean$region_category,
  drop = TRUE
)

folds <- createFolds(model_train_clean$strata, k = 5, list = TRUE)
```

```{r}
model_train_clean <- model_train_clean %>%
  select(-strata)
```

## method CV then repeatedcv
```{r}
train_control <- trainControl(
  method = "cv",
  number = 5,
  index = folds,
  classProbs = TRUE,
  summaryFunction = prSummary,  
  savePredictions = "final",
  verboseIter = TRUE
)
```

# random forest Train

```{r}
set.seed(123)
tune_random_forest_model <- train(
  high_tornado_hazard_next_year_factor ~ .,
  data = model_train_clean,
  method = "rf",
  trControl = train_control,
  tuneLength = 5,
  metric = "PR_AUC"
)

print(tune_random_forest_model)
```

```{r}
tune_random_forest_model$bestTune
```

#XG Boost Train

```{r}
set.seed(123)

  tune_xgboost_model <- train(
    high_tornado_hazard_next_year_factor ~ .,
    data = model_train_clean,
    method = "xgbTree",
    trControl = train_control,
    tuneLength = 5,
    metric = "PR_AUC"
  )

print(tune_xgboost_model)
```

## Logistic Regression Train 

```{r}
set.seed(123)
log_model <- train(
  high_tornado_hazard_next_year_factor ~ .,
  data = model_train_clean,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "PR_AUC"
)
```

#remove NAs from model_test

```{r}
model_test_clean <- model_test %>% na.omit()
```

#### random Forest evaluation ###

```{r}
random_forest_probs <- predict(tune_random_forest_model, model_test_clean, type = "prob")[, "HighRisk"]

random_forest_preds <- get_predictions(random_forest_probs)
```

## random forest: Recall for sensitivity

```{r}
confusionMatrix(random_forest_preds, 
                model_test_clean$high_tornado_hazard_next_year_factor,
                positive = "HighRisk")
```

## Storing recall for random forest 

```{r}

rf_recall <- confusionMatrix(
  random_forest_preds,
  model_test_clean$high_tornado_hazard_next_year_factor,
  positive = "HighRisk"
)$byClass["Sensitivity"]
rf_recall
```

## random forest: PR-AUC

```{r}
pr <- pr.curve(
  scores.class0 = random_forest_probs,
  weights.class0 = ifelse(model_test_clean$high_tornado_hazard_next_year_factor == "HighRisk", 1, 0)
)

pr$auc.integral
```

### XG Boost Evaluation ### 

```{r}
xgboost_probs <- predict(tune_xgboost_model, model_test_clean, type = "prob")[, "HighRisk"]

xgboost_preds <- get_predictions(xgboost_probs)
```

## XG Boost: Recall for sensitivity 

```{r}
confusionMatrix(
  xgboost_preds,
  model_test_clean$high_tornado_hazard_next_year_factor,
  positive = "HighRisk"
)
```

## Storing recall for XG Boost 

```{r}
xgb_recall <- confusionMatrix(
  xgboost_preds,
  model_test_clean$high_tornado_hazard_next_year_factor,
  positive = "HighRisk"
)$byClass["Sensitivity"]

xgb_recall
```

## XG Boost: PR AUC

```{r}
pr_xgb <- pr.curve(
  scores.class0 = xgboost_probs,
  weights.class0 = ifelse(model_test_clean$high_tornado_hazard_next_year_factor == "HighRisk", 1, 0)
)

pr_xgb$auc.integral
```

### Logistic Regression Evaluation ### 

```{r}
log_probs <- predict(log_model, model_test_clean, type = "prob")[, "HighRisk"]

log_preds <- get_predictions(log_probs)
```


```{r}
confusionMatrix(
  log_preds,
  model_test_clean$high_tornado_hazard_next_year_factor,
  positive = "HighRisk"
)
```
## Logistic Regression: Recall for sensitivity 

```{r}
log_recall <- confusionMatrix(
  log_preds,
  model_test_clean$high_tornado_hazard_next_year_factor,
  positive = "HighRisk"
)$byClass["Sensitivity"]
log_recall
```

## Logistic Regression: PR-AUC

```{r}
pr_log <- pr.curve(
  scores.class0 = log_probs,
  weights.class0 = ifelse(model_test_clean$high_tornado_hazard_next_year_factor == "HighRisk", 1, 0)
)

log_prauc <- pr_log$auc.integral
log_prauc
```

#### XG BOOST PERFORMED BEST/MOST BALANCED ###

## Reapeated k-fold 

```{r}
train_control_repeated <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 3,
  classProbs = TRUE,
  summaryFunction = prSummary,
  savePredictions = "final",
  verboseIter = TRUE
)

```

## Random Forest 

```{r}
set.seed(123)
rf_repeated <- train(
  high_tornado_hazard_next_year_factor ~ .,
  data = model_train_clean,
  method = "rf",
  trControl = train_control_repeated,
  tuneLength = 5,
  metric = "PR_AUC"
)

```

## XGBoost 

```{r}
set.seed(123)
xgb_repeated <- train(
  high_tornado_hazard_next_year_factor ~ .,
  data = model_train_clean,
  method = "xgbTree",
  trControl = train_control_repeated,
  tuneLength = 5,
  metric = "PR_AUC"
)


```


## Logistic regression 

```{r}
set.seed(123)
log_repeated <- train(
  high_tornado_hazard_next_year_factor ~ .,
  data = model_train_clean,
  method = "glm",
  family = "binomial",
  trControl = train_control_repeated,
  metric = "PR_AUC"
)
```

```{r}
evaluate_model <- function(model) {
  
  preds <- model$pred
  
  preds$pred_class <- get_predictions(preds$HighRisk)
  
  recall <- confusionMatrix(
    preds$pred_class,
    preds$obs,
    positive = "HighRisk"
  )$byClass["Sensitivity"]
  
  labels <- ifelse(preds$obs == "HighRisk", 1, 0)
  
  pr <- PRROC::pr.curve(
    scores.class0 = preds$HighRisk[labels == 1],
    scores.class1 = preds$HighRisk[labels == 0],
    curve = FALSE
  )
  
  pr_auc <- pr$auc.integral
  
  return(data.frame(
    Recall = recall,
    PR_AUC = pr_auc
  ))
}

```

```{r}

rf_results  <- evaluate_model(rf_repeated)
xgb_results <- evaluate_model(xgb_repeated)
log_results <- evaluate_model(log_repeated)
final_results <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "XGBoost"),
  Recall = c(log_results$Recall, rf_results$Recall, xgb_results$Recall),
  PR_AUC = c(log_results$PR_AUC, rf_results$PR_AUC, xgb_results$PR_AUC)
)

final_results
```


################################
###### POST HOC ANALYSIS #######
################################




## XG Boost Analysis 
## Filtering for high-risk predictions & if counted as high-risk more than once (consistently high-risk)
```{r}
model_test_full_clean <- model_test_full %>%
  filter(complete.cases(
    population_estimate,
    housing_unit_estimate,
    medvalue_estimate,
    region_category,
    avg_length,
    avg_width,
    high_tornado_hazard_next_year_factor
  ))

high_risk_predictions <- model_test_full_clean %>%
  mutate(
    predicted_class = xgboost_preds,
    predicted_prob = xgboost_probs
  ) %>%
  filter(predicted_class == "HighRisk") %>%
  arrange(desc(predicted_prob))

high_risk_predictions
```

## Numner of times a county has flagged as high-risk 
## High-Risk predictions 

```{r}
high_risk_predictions %>%
  count(county_fips, county_name_state) %>%
  filter(n > 1) %>%
  arrange(desc(n))

high_risk_predictions %>%
  count(county_fips, region_category) %>%
  filter(n > 1) %>%
  count(region_category)

```

## Most consistent high-risk counties vs. highest probability counties

```{r}
high_risk_predictions %>%
  mutate(region_name = case_when(
    region_category == "D" ~ "Dixie Alley",
    region_category == "T" ~ "Tornado Alley",
    TRUE ~ "Other"
  )) %>%
  group_by(county_fips, county_name_state, region_name) %>%
  summarise(
    times_flagged = n(),
    avg_prob = mean(predicted_prob),
    .groups = "drop"
  ) %>%
  arrange(desc(times_flagged), desc(avg_prob)) %>%
  slice_head(n = 1000)

write.csv(high_risk_predictions, "revised_high_risk_predictions.csv", row.names = FALSE)

```
## Percentage of baseline population within the cleaned model test set vs. the regions with high risk predictions for tornado hazard 

```{r}

model_test_full_clean %>%
  count(region_category) %>%
  mutate(percent = n / sum(n) * 100)

high_risk_predictions %>%
  count(region_category) %>%
  mutate(percent = n / sum(n) * 100)
```
## Confirming these are the same length 

```{r}
nrow(model_test_full_clean)
length(xgboost_preds)
```

```{r}
comparison <- data.frame(
  actual = model_test_full_clean$high_tornado_hazard_next_year_factor,
  predicted = xgboost_preds
)
comparison
```


## Missed High Risk
```{r}
missed_high_risk <- model_test_full_clean %>%
  mutate(predicted = xgboost_preds) %>%
  filter(predicted == "NoRisk" & high_tornado_hazard_next_year_factor == "HighRisk")
missed_high_risk

missed_high_risk %>%
  count(region_category)
```

```{r}
high_risk_predictions %>%
  count(region_category)
```

## Post Hoc Analysis 

```{r}
ggplot(model_test_full_clean, aes(x = xgboost_probs, 
                                 fill = high_tornado_hazard_next_year_factor)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Predicted Probability Distribution by Actual Class",
    x = "Predicted Probability",
    fill = "Actual Class"
  ) +
  scale_fill_manual(values = c(
    "NoRisk" = "red",
    "HighRisk" = "darkblue"
  )) +
  theme(
  plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  plot.subtitle = element_text(hjust = 0.5, size = 12))
```


## Plot of: Top High-Risk Counties 
```{r}
top_counties <- high_risk_predictions %>%
  mutate(region_name = case_when(
    region_category == "D" ~ "Dixie Alley",
    region_category == "T" ~ "Tornado Alley",
    TRUE ~ "Other"
  )) %>%
  group_by(county_name_state, region_name) %>%
  summarise(
    times_flagged = n(),
    avg_prob = mean(predicted_prob),
    .groups = "drop"
  ) %>%
  arrange(desc(times_flagged), desc(avg_prob)) %>%
  slice_head(n = 15)

ggplot(top_counties, aes(x = reorder(county_name_state, times_flagged), 
                         y = times_flagged, fill = region_name)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
    "Dixie Alley" = "red",
    "Tornado Alley" = "darkblue",
    "Other" = "darkgrey"
  )) +
  labs(
    title = "Top High-Risk Counties",
    subtitle = "Majority located in Dixie Alley",
    x = "County",
    y = "Times Flagged High-Risk",
    fill = "Region"
  ) +
  theme(
  plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  plot.subtitle = element_text(hjust = 0.5, size = 12))

```
## Alternate View: Top High-Risk Counties 
```{r}
top_counties <- high_risk_predictions %>%
  mutate(
    region_name = case_when(
      region_category == "D" ~ "Dixie Alley",
      region_category == "T" ~ "Tornado Alley",
      TRUE ~ "Other"
    )
  ) %>%
  group_by(county_fips, county_name_state, region_name) %>%
  summarise(
    times_flagged = n(),
    avg_prob = mean(predicted_prob),
    .groups = "drop"
  ) %>%
  arrange(desc(times_flagged), desc(avg_prob)) %>%
  slice_head(n = 10) %>%
  mutate(
    county_short = str_remove(county_name_state, ",.*")
  )



ggplot(top_counties, aes(
  x = reorder(county_short, times_flagged),
  y = times_flagged,
  fill = region_name
)) +
  geom_col() +
  coord_flip() +
  geom_text(aes(label = times_flagged), hjust = -0.1, size = 3) +
  scale_fill_manual(values = c(
    "Dixie Alley" = "red",
    "Tornado Alley" = "darkblue",
    "Other" = "grey80"
  )) +
  labs(
    title = "Top High-Risk Counties",
    subtitle = "Majority located in Dixie Alley",
    x = "County",
    y = "Times Flagged High-Risk",
    fill = "Region"
  ) +
  theme_minimal() +
  expand_limits(y = max(top_counties$times_flagged) + 1) +
  theme(
  plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  plot.subtitle = element_text(hjust = 0.5, size = 12))

```
## Plot comparing Baseline to Predicted High Risk Counties from the test set

```{r}
baseline <- model_test_full_clean %>%
  mutate(region_name = case_when(
    region_category == "D" ~ "Dixie Alley",
    region_category == "T" ~ "Tornado Alley",
    TRUE ~ "Other"
  )) %>%
  count(region_name) %>%
  mutate(percent = n / sum(n) * 100,
         type = "Baseline")

highrisk <- high_risk_predictions %>%
  mutate(region_name = case_when(
    region_category == "D" ~ "Dixie Alley",
    region_category == "T" ~ "Tornado Alley",
    TRUE ~ "Other"
  )) %>%
  count(region_name) %>%
  mutate(percent = n / sum(n) * 100,
         type = "High Risk")

combined <- bind_rows(baseline, highrisk)

ggplot(combined, aes(x = region_name, y = percent, fill = type)) +
  geom_col(position = "dodge") +
  labs(
    title = "High-Risk Tornado Predictions by Region",
    subtitle = "Comparison of baseline distribution vs predicted high-risk counties",
    x = "Region",
    y = "Percentage of Counties",
    fill = ""
  ) +
  theme_minimal()+
  scale_fill_manual(values = c(
  "Baseline" = "grey70",
  "High Risk" = "darkblue"
))+
  theme(
  plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
  plot.subtitle = element_text(hjust = 0.5, size = 12)
)
```

## Quick comparison looking at population v. median home value Dixie Alley compared to Tornado Alley. Plot to show that counties classified as higher hazard (top 25% mediam home value by region)


```{r}
county_year_hazard <- county_year_hazard %>%
  mutate(
    high_exposure = medvalue_estimate >= quantile(medvalue_estimate, 0.75, na.rm = TRUE)
  )
```
```{r}
high_risk_predictions %>%
  group_by(region_category) %>%
  summarise(
    avg_population = mean(population_estimate, na.rm = TRUE),
    avg_home_value = mean(medvalue_estimate, na.rm = TRUE)
  )

high_exposure_summary <- county_year_hazard %>%
  filter(region_category %in% c("D", "T")) %>%
  mutate(region_name = case_when(
    region_category == "D" ~ "Dixie Alley",
    region_category == "T" ~ "Tornado Alley"
  )) %>%
  count(region_name, high_exposure) %>%
  group_by(region_name) %>%
  mutate(percent = n / sum(n) * 100)

ggplot(high_exposure_summary, aes(x = region_name, y = percent, fill = high_exposure)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
    "TRUE" = "darkblue",
    "FALSE" = "grey70"
  ), labels = c("Not High Exposure", "High Exposure")) +
  labs(
    title = "Share of High-Exposure Counties by Region",
    subtitle = "Based on top 25% of median home value",
    x = "Region",
    y = "Percentage of Counties",
    fill = "Exposure Level"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12)
  )
```

